This thesis is only the tip of a six-year-long journey, which I was privileged enough to share with many wonderful people whose contribution to my success and wellbeing I cannot appreciate enough.
Like most long journeys, not every part of it was pleasant (although many were), and at times I came close to losing hope.
My supervisor Stefan Brunthaler, however, never lost hope in me and believed in me even when I could barely do so myself.
I am immensely grateful to Stefan, not only for his unwavering belief, but also for his patient guidance, his hands-on support and his readiness to have his ideas challenged.
I cherish our countless technical debates and arguments, usually over an even less countable number of beers, every bit as much as our collaboration itself.

Special thanks belong to Vasil Sarafov, who apart from being an excellent mental sparring partner, also took it upon himself to act as a \enquote{remote hand} on several occasions, saving me many trips to Munich.
The number of beers I owe him for these many favors likely exceeds what he can (or should) drink in the forthcoming years.
I am also grateful to Matthias Bernad, whom I am lucky enough to count as both a close friend and a colleague and who proved to be as good at the latter as he is at the former.
I want to thank David Markvica and Alexander Diem for the many insightful discussions and for keeping the \enquote{Wiener Schmäh} alive.
I owe thanks to Gergö Barany and Matthias Neugschwandtner who taught me a tremendous amount during my internships at Oracle.
Together with Florian Schwarcz, they also made co-authorship a genuine pleasure.

Finally, I am deeply grateful to my family, without whom this entire endeavor would never have been possible.
My wonderful wife Christina, who stood by me in good times and in bad and who had to sacrifice plenty to support me over all this time.
And my parents, Geraldine and Johann, who not only provided all the support one could hope for, but also fostered a genuine curiosity in the world for as long as I can remember, which ultimately culminated in this thesis.
I dedicate this thesis to my lovely daughter Isabella, who was born during my PhD and brings me joy every single day.
I can only hope to instill the same curiosity in her that was gifted to me and for her to have such wonderful companions.
